---
title: "DisCo ParTI Polytopes for T1D Data"
author: "Duy An Le, Sean Pascoe"
date: today
format: 
  html:
    embed-resources: true
    toc: true
    link-external-newwindow: true
    df-print: paged
    footnotes-hover: true
title-block-banner-color: black
execute:
  echo: false
  warning: false
from: markdown+emoji
editor_options: 
  chunk_output_type: console
---
::: {.callout-tip icon="false"}
## Github Repo Link

<https://github.com/pascoesean/discoparti>
:::

## Overview

Welcome to our polytope document!

Here you can view each polytope from each cell cluster identified by leiden clustering on the
input data. Each polytope is defined in $n-1$ dimensional space, where $n$ is the number of
vertices (archetypes) of that shape. For ease of viewing, all polytopes are visualized with
the first 3 princpal components that define that $n$ dimensional space. Polytope fitting code was
adapted from the [Uri Alon Lab](https://www.weizmann.ac.il/mcb/alon/download/pareto-task-inference-parti-method) [Korem et al., 2015](http://www.weizmann.ac.il/mcb/UriAlon/sites/mcb.UriAlon/files/korem_2015_-_geometry_of_the_gene_expression_space.pdf).

Mousing over an archetype will give you information about the top 15 genes that were highly
weighted at that point in space compared to the centroid of the polytope, as well as gene set
enrichment done on the population of cells nearest to that polytope. We also provide a normalized
phenotype enrichment score that ranges between -1 and 1 for association with T1D, as well as the 
z score of this score in the distribution of of enrichment scores that appear with shuffled labels
at the same points in space. Additional information is available in the Appendix.


```{r}
# load packages ----
source('parti_output_wrangling.R')

```

## PANCREATIC ISLETS

```{r}
## ACINAR CELLS --- 
pc_filepath = "parti_outputs/AcinarCells/lite_pc.csv"
labels_filepath = "parti_inputs/islets/Acinar Cells/cell_labels.csv"
enrichment_filepath = "parti_outputs/AcinarCells/lite_continuous_significant.csv"
gene_names_filepath = "parti_inputs/islets/Acinar Cells/gene_names.csv"
arch_orig_filepath = "parti_outputs/AcinarCells/lite_arcOrig.csv"
arch_filepath = "parti_outputs/AcinarCells/lite_arc.csv"
pea_filepath = 'pea_outputs/AcinarCells/pea_output.csv'
title = 'Acinar Cells'

plot_parti(pc_filepath, labels_filepath, enrichment_filepath,
           gene_names_filepath, arch_orig_filepath, arch_filepath, pea_filepath, title)
```


```{r}
#| label: alpha cells

## ALPHA CELLS --- 
pc_filepath = "parti_outputs/AlphaCells/lite_pc.csv"
labels_filepath = "parti_inputs/islets/Alpha Cells/cell_labels.csv"
enrichment_filepath = "parti_outputs/AlphaCells/lite_continuous_significant.csv"
gene_names_filepath = "parti_inputs/islets/Alpha Cells/gene_names.csv"
arch_orig_filepath = "parti_outputs/AlphaCells/lite_arcOrig.csv"
arch_filepath = "parti_outputs/AlphaCells/lite_arc.csv"
pea_filepath = 'pea_outputs/AlphaCells/pea_output.csv'
title = 'Alpha Cells'


plot_parti(pc_filepath, labels_filepath, enrichment_filepath,
           gene_names_filepath, arch_orig_filepath, arch_filepath, pea_filepath, title)
```


```{r}
#| label: alpha cells ghi

## ALPHA ghi CELLS --- 
pc_filepath = "parti_outputs/AlphaCellsGHi/lite_pc.csv"
labels_filepath = "parti_inputs/islets/Alpha Cells G_hi/cell_labels.csv"
enrichment_filepath = "parti_outputs/AlphaCellsGHi/lite_continuous_significant.csv"
gene_names_filepath = "parti_inputs/islets/Alpha Cells G_hi/gene_names.csv"
arch_orig_filepath = "parti_outputs/AlphaCellsGHi/lite_arcOrig.csv"
arch_filepath = "parti_outputs/AlphaCellsGHi/lite_arc.csv"
pea_filepath = 'pea_outputs/AlphaCellsGHi/pea_output.csv'
title = 'Alpha Cells, High Glucagon'


plot_parti(pc_filepath, labels_filepath, enrichment_filepath,
           gene_names_filepath, arch_orig_filepath, arch_filepath, pea_filepath, title)
```

```{r}


## BETA CELLS --- 
pc_filepath = "parti_outputs/BetaCells/full_pc.csv"
labels_filepath = "parti_inputs/islets/Beta Cells/cell_labels.csv"
enrichment_filepath = "parti_outputs/BetaCells/full_continuous_significant.csv"
gene_names_filepath = "parti_inputs/islets/Beta Cells/gene_names.csv"
arch_orig_filepath = "parti_outputs/BetaCells/full_arcOrig.csv"
arch_filepath = "parti_outputs/BetaCells/full_arc.csv"
pea_filepath = 'pea_outputs/BetaCells/pea_output.csv'
title = 'Beta Cells'


plot_parti(pc_filepath, labels_filepath, enrichment_filepath,
           gene_names_filepath, arch_orig_filepath, arch_filepath, pea_filepath, title)
```


```{r}
#| label: Delta cells

## DELTA CELLS --- 
pc_filepath = "parti_outputs/DeltaCells/full_pc.csv"
labels_filepath = "parti_inputs/islets/Delta Cells/cell_labels.csv"
enrichment_filepath = "parti_outputs/DeltaCells/full_continuous_significant.csv"
gene_names_filepath = "parti_inputs/islets/Delta Cells/gene_names.csv"
arch_orig_filepath = "parti_outputs/DeltaCells/full_arcOrig.csv"
arch_filepath = "parti_outputs/DeltaCells/full_arc.csv"
pea_filepath = "pea_outputs/DeltaCells/pea_output.csv"
title = 'Delta Cells'


plot_parti(pc_filepath, labels_filepath, enrichment_filepath,
           gene_names_filepath, arch_orig_filepath, arch_filepath, pea_filepath, title)

```


```{r}
#| label: Ductal cells

## DUCTAL CELLS --- 
pc_filepath = "parti_outputs/DuctalCells/lite_pc.csv"
labels_filepath = "parti_inputs/islets/Ductal Cells/cell_labels.csv"
enrichment_filepath = "parti_outputs/DuctalCells/lite_continuous_significant.csv"
gene_names_filepath = "parti_inputs/islets/Ductal Cells/gene_names.csv"
arch_orig_filepath = "parti_outputs/DuctalCells/lite_arcOrig.csv"
arch_filepath = "parti_outputs/DuctalCells/lite_arc.csv"
pea_filepath = 'pea_outputs/DuctalCells/pea_output.csv'
title = 'Ductal Cells'


plot_parti(pc_filepath, labels_filepath, enrichment_filepath,
           gene_names_filepath, arch_orig_filepath, arch_filepath, pea_filepath, title)
```


```{r}
#| label: Endothelial cells

## Endothelial CELLS --- 
pc_filepath = "parti_outputs/EndothelialCells/lite_pc.csv"
labels_filepath = "parti_inputs/islets/Endothelial Cells/cell_labels.csv"
enrichment_filepath = "parti_outputs/EndothelialCells/lite_continuous_significant.csv"
gene_names_filepath = "parti_inputs/islets/Endothelial Cells/gene_names.csv"
arch_orig_filepath = "parti_outputs/EndothelialCells/lite_arcOrig.csv"
arch_filepath = "parti_outputs/EndothelialCells/lite_arc.csv"
pea_filepath = 'pea_outputs/EndothelialCells/pea_output.csv'
title = 'Endothelial Cells'


plot_parti(pc_filepath, labels_filepath, enrichment_filepath,
           gene_names_filepath, arch_orig_filepath, arch_filepath, pea_filepath, title)
```

## PBMCs

```{r}
#| label: B cells

## B CELLS --- 
pc_filepath = "parti_outputs/BCells/lite_pc.csv"
labels_filepath = "parti_inputs/pbmcs/B Cells/cell_labels.csv"
enrichment_filepath = "parti_outputs/BCells/lite_continuous_significant.csv"
gene_names_filepath = "parti_inputs/pbmcs/B Cells/gene_names.csv"
arch_orig_filepath = "parti_outputs/BCells/lite_arcOrig.csv"
arch_filepath = "parti_outputs/BCells/lite_arc.csv"
pea_filepath = 'pea_outputs/BCells/pea_output.csv'
title = 'B Cells'


plot_parti(pc_filepath, labels_filepath, enrichment_filepath,
           gene_names_filepath, arch_orig_filepath, arch_filepath, pea_filepath, title)
```

```{r}
#| label: CD14+ monocytes

## CD14+ MONOCYTES --- 
pc_filepath = "parti_outputs/CD14+Monocytes/lite_pc.csv"
labels_filepath = "parti_inputs/pbmcs/CD14+ monocytes/cell_labels.csv"
enrichment_filepath = "parti_outputs/CD14+Monocytes/lite_continuous_significant.csv"
gene_names_filepath = "parti_inputs/pbmcs/CD14+ monocytes/gene_names.csv"
arch_orig_filepath = "parti_outputs/CD14+Monocytes/lite_arcOrig.csv"
arch_filepath = "parti_outputs/CD14+Monocytes/lite_arc.csv"
pea_filepath = 'pea_outputs/CD14+Monocytes/pea_output.csv'
title = 'CD14+ Monocytes'


plot_parti(pc_filepath, labels_filepath, enrichment_filepath,
           gene_names_filepath, arch_orig_filepath, arch_filepath, pea_filepath, title)
```

```{r}
#| label: CD16+ monocytes

## CD16+ MONOCYTES --- 
pc_filepath = "parti_outputs/CD16+Monocytes/lite_pc.csv"
labels_filepath = "parti_inputs/pbmcs/CD16+ monocytes/cell_labels.csv"
enrichment_filepath = "parti_outputs/CD16+Monocytes/lite_continuous_significant.csv"
gene_names_filepath = "parti_inputs/pbmcs/CD16+ monocytes/gene_names.csv"
arch_orig_filepath = "parti_outputs/CD16+Monocytes/lite_arcOrig.csv"
arch_filepath = "parti_outputs/CD16+Monocytes/lite_arc.csv"
pea_filepath = 'pea_outputs/CD16+Monocytes/pea_output.csv'
title = 'CD16+ Monocytes'


plot_parti(pc_filepath, labels_filepath, enrichment_filepath,
           gene_names_filepath, arch_orig_filepath, arch_filepath, pea_filepath, title)
```

```{r}
#| label: Dendritic cells

## T CELLS --- 
pc_filepath = "parti_outputs/DendriticCells/lite_pc.csv"
labels_filepath = "parti_inputs/pbmcs/DCs/cell_labels.csv"
enrichment_filepath = "parti_outputs/DendriticCells/lite_continuous_significant.csv"
gene_names_filepath = "parti_inputs/pbmcs/DCs/gene_names.csv"
arch_orig_filepath = "parti_outputs/DendriticCells/lite_arcOrig.csv"
arch_filepath = "parti_outputs/DendriticCells/lite_arc.csv"
pea_filepath = 'pea_outputs/DendriticCells/pea_output.csv'
title = 'Dendritic Cells'


plot_parti(pc_filepath, labels_filepath, enrichment_filepath,
           gene_names_filepath, arch_orig_filepath, arch_filepath, pea_filepath, title)
```

```{r}
#| label: ISG hi T cells

## T CELLS --- 
pc_filepath = "parti_outputs/ISGhiTCells/lite_pc.csv"
labels_filepath = "parti_inputs/pbmcs/ISGhi T Cells/cell_labels.csv"
enrichment_filepath = "parti_outputs/ISGhiTCells/lite_continuous_significant.csv"
gene_names_filepath = "parti_inputs/pbmcs/ISGhi T Cells/gene_names.csv"
arch_orig_filepath = "parti_outputs/ISGhiTCells/lite_arcOrig.csv"
arch_filepath = "parti_outputs/ISGhiTCells/lite_arc.csv"
pea_filepath = 'pea_outputs/ISGhiTCells/pea_output.csv'
title = 'T Cells, high IFN signaling'


plot_parti(pc_filepath, labels_filepath, enrichment_filepath,
           gene_names_filepath, arch_orig_filepath, arch_filepath, pea_filepath, title)
```

```{r}
#| label: NK cells

## T CELLS --- 
pc_filepath = "parti_outputs/NKCells/lite_pc.csv"
labels_filepath = "parti_inputs/pbmcs/NK cells/cell_labels.csv"
enrichment_filepath = "parti_outputs/NKCells/lite_continuous_significant.csv"
gene_names_filepath = "parti_inputs/pbmcs/NK cells/gene_names.csv"
arch_orig_filepath = "parti_outputs/NKCells/lite_arcOrig.csv"
arch_filepath = "parti_outputs/NKCells/lite_arc.csv"
pea_filepath = 'pea_outputs/NKCells/pea_output.csv'
title = 'NK Cells'


plot_parti(pc_filepath, labels_filepath, enrichment_filepath,
           gene_names_filepath, arch_orig_filepath, arch_filepath, pea_filepath, title)
```

```{r}
#| label: NKT cells

## T CELLS --- 
pc_filepath = "parti_outputs/NKTCells/lite_pc.csv"
labels_filepath = "parti_inputs/pbmcs/NKT Cells/cell_labels.csv"
enrichment_filepath = "parti_outputs/NKTCells/lite_continuous_significant.csv"
gene_names_filepath = "parti_inputs/pbmcs/NKT Cells/gene_names.csv"
arch_orig_filepath = "parti_outputs/NKTCells/lite_arcOrig.csv"
arch_filepath = "parti_outputs/NKTCells/lite_arc.csv"
pea_filepath = 'pea_outputs/NKTCells/pea_output.csv'
title = 'NKT Cells'


plot_parti(pc_filepath, labels_filepath, enrichment_filepath,
           gene_names_filepath, arch_orig_filepath, arch_filepath, pea_filepath, title)
```

```{r}
#| label: t cells

## T CELLS --- 
pc_filepath = "parti_outputs/TCells/lite_pc.csv"
labels_filepath = "parti_inputs/pbmcs/T cells/cell_labels.csv"
enrichment_filepath = "parti_outputs/TCells/lite_continuous_significant.csv"
gene_names_filepath = "parti_inputs/pbmcs/T cells/gene_names.csv"
arch_orig_filepath = "parti_outputs/TCells/lite_arcOrig.csv"
arch_filepath = "parti_outputs/TCells/lite_arc.csv"
pea_filepath = 'pea_outputs/TCells/pea_output.csv'
title = 'T Cells'


plot_parti(pc_filepath, labels_filepath, enrichment_filepath,
           gene_names_filepath, arch_orig_filepath, arch_filepath, pea_filepath, title)
```


## Appendix

### Data Sources

- Pancreatic islet data was downloaded from [PANC-DB](https://hpap.pmacs.upenn.edu/). 22 Adults
were chosen to create a cohort of age and sex matched healthy controls.

- PBMC data was derived from [Honardoost et al., 2024](https://genomemedicine.biomedcentral.com/articles/10.1186/s13073-024-01300-z), and downloaded from [synapse.org](https://www.synapse.org/Synapse:syn53641849).

### Phenotype Enrichment Analysis

To calculate the enrichment of each phenotype at each archetype throughout the reduced dimensional
space that the polytopes are defined in, we derived a normalized enrichment score as the max
deviation from zero calculated from a running score based on euclidean distance of each point (cell) to each archetype. For each archetype, the process was as follows:

1. Rank all cells from shortest to longest euclidean distance to that archetype

2. For each cell, add $s_{hit}$ to the running score if the cell is in the phenotype of interest. Otherwise, subtract $s_{miss}$, where $s_{hit}$ is the absolute value of the mean-centered distance to that archetype scaled by the total score of observations in that phenotype, and $s_{miss}$ is one over the number of observations not in that phenotype, i.e.

$$ s_{hit} = \frac{|d_{i}|}{D_p} \text{, where } D_p = \sum_{i=1}^{p}{|d_i|}$$
$$ s_{miss} = \frac{1}{N_{\neg p}}$$
Here, $D_p$ is the summed absolute value of the mean centered distances for all $p$ values in the phenotype of interest, and $N_\neg p$ is the number of observations that are not in the phenotype of interest.


3. Report maximum deviation from zero of the running score as the normalized enrichment score.
